---
title: "Statistical Data Analysis of Student Goals"
author: Mateusz Zaremba
date: November 9, 2019
output: 
  pdf_document:
    toc: false
    toc_depth: 2
    number_sections: true
    fig_caption: false
    df_print: kable
    highlight: tango
    citation_package: natbib
    keep_tex: false
fontzise: 11pt
geometry: margin=1.5in
documentclass: report
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE, error = FALSE, tidy = FALSE, cache = FALSE, fig.cap = "...")
```

```{r}
# load external script
source("source/DataManipulation.R")
```

# Abstract
This paper analyses the data gathered from surveying 625 undergraduate students. The authors of the survey tried to prove two hypothesis: 1) During students' junior years, they tend to primarily focus on getting good grades while during their senior years, the focus shifts towards a deep-understanding of the subject and 2) students' enjoyment and interest tends to deteriorate as they progress through their studies. It is not obvious why this might be the case and if the student's sex or studied subject has any bearing. This is why the survey has 15 questions and probes 7 assessment categories. Each category consists of 3 to 1 question and because the order of the questions is randomised, the student should not know the categories nor notice any patterns.

The data manipulation was done using R and tidywerse packages. A full analysis will be presented, including data: preparation, analysis, exploration and interpretation; calculation of confidence interval for a proportion, interpretation of the results using different kinds of graphs and an explanation of the methods used.

# Introduction
It is interesting how undergraduate students' goals change through-out their studies. They often experience various syndromes like burnout, impostor, disheartening or even attempt a suicide. A Harvard graduate, Alex Chang, in his TEDx talk titled ["The Unspoken Reality Behind the Harvard Gates"](https://www.youtube.com/watch?v=WzP7oDCciGI) speaks about the pressure of getting the best grades; how he was called for a jasmine tea to his tutor and asked if he couldn't give it his all, while he already was doing the best he could. He also recalls one tragic night when he and his roommates were woken up at 4am, to be informed that one of his friends has taken his own life.

Because this paper is going to be talking about student's course enjoyment, expectations and his or her focus on grades vs. understanding I would like to give it another, less visible shade for there might be a lot more to say about a student who is at the bottom of the scale. It was assumed that a student, who might be at risk of developing mental health problems, would be someone who: is not enjoying the course, finds it not interesting but still primarily aims to perform better than others, and is led by the fear of performing poorly. We will try to identify such students, calculate the confidence-interval-for-a-proportion of finding them, and test the hypotheses.

# Data
The data we will be analysing was originally sourced from *Elliot, A. J. and McGregor, H. A. (2001). A 2 x 2 achievement-goal framework. Journal of Personality and Social Psychology, 80, 3, 501-519*.

## Collection
The data was already collected but my own survey results were added to the data set.

## Initialising
The data was converted from the original `.xlsx` format to `.csv` using *Microsoft Excel for Mac* and then it was loaded to *R* script using the *tidyverse* package - [*readr*](https://readr.tidyverse.org/index.html]).

The initial number of students was also saved in a variable for later calculations.

This is how the data looked like after loading it into the *R* script and before cleaning:

```{r}
studentGoalsEg <- createExampleTable(StudentGoalsData, 6, 9, 1, 6, 16, 19)
studentGoalsEg
```

## Cleansing
First, the *seq* column was dropped since it does not serve any purpose. Second, rows with empty cells were dropped because they could falsify the results.

## Coding
The following coding information was applied to the data:

| Subject | Sex | Code |
| ------- | ---- | --- |
| Management | Male | 1 |
| Law | Female | 2
| Tourism | - | 3 |
| General Economics  | - | 4 |
| Accounting  | - | 5 |
| Statistics | - | 6 |

E.g., the code for *Male* was *1*, so the cells in the *sex* column containing *1* were replaced with a *Male* string; the code for *General Economics* was *4*, so the cells in the *subject* column containing *4* were replaced with a *General Economics* string. This is how the *sex* and *subject* columns looked like after applying the coding information:

```{r}
mean_dat[6:9,3:4]
```

## Assessment Categories
The survey's instructions provided 7 assessment categories and their appropriate questions. You can see them below with added labelling and interpretation.

Category | Label | Interpretation | Question |
-------- | ----- | -------------- | -------- |
Performance Approach | M1 | Importance of doing better than others |1, 2, 3 |
Performance Avoidance | M2 | Motivation based on the fear of performing poorly | 4, 5, 6 |
Mastery Approach | M3 | Students' grade-orientation focus | 7, 8, 9 |
Mastery Avoidance | M4 | Students' fear of not mastering the course | 10, 11, 12 |
Interest | IR | Student expects the course to be interesting | 13 |
Enjoyment | EJ | Student expects the course to be enjoyable | 14 |
Importance focus | MG | Students' importance focus on understanding vs. grades | 15 |
Table: Interpretation table

## De-randomization and Renaming
Random order (6, 12, 11, 1, 7, 2, 10, 8, 5, 3, 9, 4) was given to use in the survey's instructions. This meant, e.g., *question 6* from the *Performance Avoidance* category is numbered *1* in the survey and in the data set; *question 12* from the *Mastery Avoidance* category is numbered *2* in the survey and in the set; and so on. 

The column naming changed in the process: ***q****1* became ***Q****6*, changing lower case ***q*** to upper case ***Q*** as well as the number of the question.

*Table 4* illustrates the entire process and assigns questions to their appropriate categories:

| |
| - | - | - | - | - | - | - | - | - | - | - | - | - |
Data set order | q1 | q2 | q3 | q4 | q5 | q6 | q7 | q8 | q9 | q10 | q11 | q12 |
Survey order | Q6 | Q12 | Q11 | Q1 | Q7 | Q2 | Q10 | Q8 | Q5 | Q3 | Q9 | Q4 |
Table: In the survey, questions from 1 to 12 were derandomized and renamed

Questions 13, 14 and 15 - which were in fact, hypotheses testing questions and *Interest*, *Enjoyment*, *Importance focus* assessment-categories accordingly - kept the same order in the survey and in the data set, so they were only renamed for easier data manipulation:

| Previous column name | New column name |
| - | - |
| interest | IG |
| enjoy | EJ |
| mastgrad | MG |

## Clean Data
This is how the data looked like after cleaning:

```{r}
meanTableEg <- createExampleTable(mean_dat, 6, 9, 1, 6, 15, 19)
meanTableEg
```

## Visual Exploration
For all the students, the means from *M1*, *M2*, *M3* and *M4* assessment categories were shown on the graphs below; data smoothing was applied in attempt to capture any important patterns.

### M1 - Performance Approach
```{r}
# M1
# Plot mean results of performance approach questions
# for all students with relation to student's year, sex and subject
# data
d <- ggplot(data = dat, aes(mean_dat$year, mean_dat$M1))
# mapping data (use "jitter" to improve the graph and avoid gridding)
l <- d + geom_jitter(aes(colour = sex, shape = subject))
# smoothing
s <- l + geom_smooth(method = loess, formula = y ~ x, se = TRUE)
# adding theme
t <- s + theme_grey()
# adding colouring
c <- t
# adding labels
c + labs(
  tag = "M1",
  title = "Performance Approach set on basis of year of study, sex and subject.",
  subtitle = "How important it is to students to do better than others?",
  x = "Year (1-4)",
  y = "Answer: 1 (Low) - 7 (High)",
  colour = "Sex",
  shape = "Subject"
)
```
It is visible that the students' performance approach almost does not change, and all the variance are within the confidence interval.

### M2 - Performance Avoidance
```{r}
# M2
# Plot mean results of performance avoidance
# for all students with relation to student's year, sex and subject
# data
d <- ggplot(data = dat, aes(year, mean_dat$M2))
# mapping data (use "jitter" to improve the graph and avoid gridding)
l <- d + geom_jitter(aes(colour = sex, shape = subject))
# smoothing
s <- l + geom_smooth(method = loess, formula = y ~ log(x), se = TRUE)
# adding theme
t <- s + theme_grey()
# adding colouring
c <- t
# adding labels
c + labs(
  tag = "M2",
  title = "Performance Avoidance set on basis of year of study, sex and subject.",
  subtitle = "How motivated are students by fear of performing poorly?",
  x = "Year: 1 - 4",
  y = "Answer: 1 (Low) - 7 (High)",
  colour = "Sex",
  shape = "Subject"
)
```
Comparing year's 1 average results to year's 4, it is clearly visible that the students' performance avoidance drops by 1 scale point. 

### M3 - Mastery Approach
```{r}
# M3
# Plot mean results of Mastery Approach questions
# for all students with relation to student's year, sex and subject
# data
d <- ggplot(data = dat, aes(year, mean_dat$M3))
# mapping data (use "jitter" to improve the graph and avoid gridding)
l <- d + geom_jitter(aes(colour = sex, shape = subject))
# smoothing
s <- l + geom_smooth(method = stats::loess, formula = y ~ log(x), se = TRUE)
# adding theme
t <- s + theme_grey()
# adding colouring
c <- t
# adding labels
c + labs(
  tag = "M3",
  title = "Mastery Approach set on basis of year of study, sex and subject.",
  subtitle = "Students' grade-orientation focus",
  x = "Year: 1 - 4",
  y = "Answer: 1 (Low) - 7 (High)",
  colour = "Sex",
  shape = "Subject"
)
```
It is visible that the student's mastery approach drops by almost 1 scale point.

### M4 - Mastery Avoidance
```{r}
# M4
# Plot mean results of Mastery Avoidance questions
# for all students with relation to student's year, sex and subject
# data
d <- ggplot(data = dat, aes(year, mean_dat$M4))
# mapping data (use "jitter" to improve the graph and avoid gridding)
l <- d + geom_jitter(aes(colour = sex, shape = subject))
# smoothing
s <- l + geom_smooth(method = stats::loess, formula = y ~ log(x), se = TRUE)
# adding theme
t <- s + theme_grey()
# adding colouring
c <- t
# adding labels
c + labs(
  tag = "M4",
  title = "Mastery Avoidance set on basis of year of study, sex and subject.",
  subtitle = "Students' fear of not mastering the course.",
  x = "Year: 1 - 4",
  y = "Answer: 1 (Low) - 7 (High)",
  colour = "Sex",
  shape = "Subject"
)
```
There is a dropping tendency for *Mastery Avoidance* but it is not drastic.

### Sex and Subject
```{r}
#subject by gender and normalizing using position = "dodge" to place overlapping objects directly beside one another
ggplot(data = dat) + 
  geom_bar(mapping = aes(x = sex, fill = subject), position = "dodge") +
  labs(
    title = "Subject student count with regards to sex"
  )
```

# Methodology
The data was manipulated using `R` programming language and `tidyverse` packages:

- [`ggplot2`](https://ggplot2.tidyverse.org/) - data visualisation
- [`dplyr`](https://dplyr.tidyverse.org/) - data manipulation
- [`tibble`](https://tibble.tidyverse.org/) - data reimagining
- [`readr`](https://readr.tidyverse.org/) - data reading
- [`tidyr`](https://tidyr.tidyverse.org/) - data cleaning
- [`purrr`](https://purrr.tidyverse.org/) - syntax simplification

Around 625 students were surveyed. They answered on a 7-level scale; 1 meaning the student felt the statement asked in the question is 'Not true of him/her' and 7 meaning the student felt it was 'Very true of him/her'. See the table below for a graphical explanation:

 | | | | | | | |
- | - | - | - | - | - | - | - | -
Not true of me | 1 | 2 | 3 | 4 | 5 | 6 | 7 | Very true of me

## Mean
Apart from *Interest*, *Enjoyment* and *Importance focus* categories, *Performance Approach*, *Performance Avoidance*, *Mastery Approach*, *Mastery Avoidance* categories consisted of 3 questions; for these 4 categories the means were computed and saved for each individual student.

This resulted in 4 extra columns added to the original data set. These results were using in data exploration and finding a student-at-risk. An example of these can be seen below:
```{r}
dat_tibble[1:5, 20:23]
```

## Confidence Interval for a Proportion
To find students at risk with 95% confidence, we will calculate confidence interval for a proportion.

### Equations
$$ \hat{p} = \frac{x}{n} = \frac{events}{trials} $$

$$ \hat{p} \pm Z \sqrt{\frac{\hat{p}(1 - \hat{p})}{n}} $$

### Calculations
Calculating a 95% confidence interval for the student-at-risk population proportion.

$$ \hat{p} = \frac{2}{625} = 0.0032 $$

$$ Z = 1.96  $$

$$ \hat{p} \pm Z = 0.0032 \pm 1.96 \sqrt{\frac{{0.0032}(1 - {0.0032})}{625}} = 0.0032 \pm 0.0044 $$

# Results
## Hypothesis Testing
To test the hypotheses the following categories where picked:

- *Hypothesis 1* - *Importance focus (MG)*
- *Hypothesis 2* - *Enjoyment (EJ)* and *Interest (IR)*

The graphs below are a visual investigation of the hypotheses using their chosen categories. All the students are represented; student's sex was made distinguishable by a colour and student's subject was made visible by a shape.

## Hypothesis 1
```{r}
# 1) During students’ junior years, they tend to primarily focus on getting good grades while during their senior years, the focus shifts towards a deep understanding of the subject.

# Plot set on basis of the student's sex.
# Plot mean results of (Primarily understanding/Equal Importance/Primarily grades)scale 
# for all students with relation to student's year, sex and subject
# data
d <- ggplot(data = dat, aes(year, mean_dat$MG))
# mapping data (use "jitter" to improve the graph and avoid gridding)
l <- d + geom_jitter(aes(colour = mean_dat$sex, shape = mean_dat$subject))
# smoothing
s <- l + geom_smooth(method = stats::loess, formula = y ~ log(x), se = TRUE)
# adding theme
t <- s + theme_grey()
# adding colouring
c <- t
# adding labels
c + labs(
  tag = "MG",
  title = "Student's importance scale between understanding and grades set on basis of:
different years of study, sexes and subjects.",
  subtitle = "Scale: Primarily understanding (1) / Equal Importance (4) / Primarily grades (7)",
  x = "Year (1-4)",
  y = "Scale: 1 (Understanding) - 4 (Equal) - 7 (Grades)",
  colour = "Sex",
  shape = "Subject"
)
```
The hypothesis stated was:

> During students’ junior years, they tend to primarily focus on getting good grades while during their senior years, the focus shifts towards a deep understanding of the subject. 

Visual investigation shows that students across all years, almost never primarily focus on grades. Most of the time they show equal interest in grades and understanding. Graph smoothing does indeed show a slight shift towards deeper understanding at the more senior years but 

Because the hypothesis wrongly assumes that students at their junior years primarily focus on getting good grades, and visual investigation clearly shows that the student's focus remains mostly equal, with a slight shift towards understanding at the senior year, the hypothesis is deemed wrong and no further investigation will be carried out.

## Hypothesis 2
Let us now remind ourselves the hypothesis 2: 

> Students’ enjoyment and interest tends to deteriorate as they progress through their studies.

### Interest
```{r}
# 2) students’ enjoyment and interest tends to deteriorate as they progress through their studies. 
# It is not obvious why this might be the case and if the student’s sex or studied subject has any bearing

# interest
# Plot set on basis of the student's sex.
# Plot mean results of course interestedness expectations questions
# for all students with relation to student's year, sex and subject
# data
d <- ggplot(data = dat, aes(year, mean_dat$IR))
# mapping data (use "jitter" to improve the graph and avoid gridding)
l <- d + geom_jitter(aes(colour = mean_dat$sex, shape = mean_dat$subject))
# smoothing
s <- l + geom_smooth(method = stats::loess, formula = y ~ log(x), se = TRUE)
# adding theme
t <- s + theme_grey()
# adding colouring
c <- t
# adding labels
c + labs(
  tag = "IR",
  title = "Student's course interestedness expectations set on basis of:
different years of study, sexes and subjects.",
  subtitle = "Student response: \'I expect my courses this semester to be very interesting\'",
  x = "Year (1-4)",
  y = "Expectations: 1 (Low) - 7 (High)",
  colour = "Sex"
)
```
Given the size of the population being quite small and graph smoothing not showing any visible patters, it is fair to say that the student's interest remains the same during their time at the university.

### Enjoyment
```{r}
# enjoy
# Plot set on basis of the student's sex.
# Plot mean results of course interestedness expectations questions
# for all students with relation to student's year, sex and subject
# data
d <- ggplot(data = dat, aes(year, mean_dat$EJ))
# mapping data (use "jitter" to improve the graph and avoid gridding)
l <- d + geom_jitter(aes(colour = mean_dat$sex, shape = mean_dat$subject))
# smoothing
s <- l + geom_smooth(method = stats::loess, formula = y ~ log(x), se = TRUE)
# adding theme
t <- s + theme_grey()
# adding colouring
c <- t
# adding labels
c + labs(
  tag = "EJ",
 title = "Student's course enjoyment expectations set on basis of:
different years of study, sexes and subjects.",
  subtitle = "Student response: \'I expect my courses this semester to be very enjoyable\'",
  x = "Year (1-4)",
  y = "Expectations: 1 (Low) - 7 (High)",
  colour = "Sex"
)
```
Enjoyment seems to follow the same patter. Because its slight downwards trend is with the confidence interval it is deemed insignificant.

It is clearly visible that neither students' interest nor enjoyment deteriorates over the course of their studies. Visual investigation showed that it remains roughly the same.

## Finding Students at Risk
To find out if there are any students at risk the arbitrary criteria was assumed:

- *Performance Approach* $\geq$ 6
- *Performance Avoidance* $\geq$ 6
- *Interest* $\leq$ 2
- *Enjoyment* $\leq$ 2
- *Importance focus* $\geq$ 4

Using this conservative criteria, 2 students were found (If we mitigated the scale only by 1 point, the number of students would go up to 7); they are not enjoying their course nor find it interesting, but their main focus remains on doing better than others and avoiding performing poorly with a main or neutral focus on grades. It was assumed that such students might be at risk of developing mental health problems.

Students at risk found using conservative criteria:
```{r}
mm2[,c("age", "sex", "subject", "M1", "M2", "IR", "EJ", "MG")]
```

Students at risk found using mitigated criteria:
```{r}
mm3[,c("age", "sex", "subject", "M1", "M2", "IR", "EJ", "MG")]
```

### Confidence Interval for a Proportion of Students at Risk
The upper confidence interval for a proportion is `r ci_up[,]`, and the lower is `r ci_low[,]`, which gives us a confidence interval, for proportion of finding a student-at-risk as: CI = (0.00762, -0.00122). This mean we could say with 95% confidence the percentage of the times we should expect to find a student at risk is between 0.7% and 0%.

# Conclusion
Looking at data, we see there are in fact 2 students who might be at risk. They are both 18-year-old females studying *General Economic* at their sophomore years. Usually when we investigate the data, the extreme cases are dropped. But this time it would mean that we would be dismissing students who might be at real risk. Of course, it might be the case that they mixed up the scale or had a bad day, but this begs the question - should such cases be dismissed? Could universities address this data in any way? We might say that it is only a tiny fraction of the population and chances of finding such a student are less than 1%. So are the number of students taking their own lives at campuses.

I believe something could be done and such data should never be dismissed because we might be able to help these students. Maybe universities could develop their own programmes, that students could use to self-diagnose, whether they are at risk of developing mental health problems and if they are, it could direct them towards a university's counsellor as well as other places. Such tests already exist and can be found on the NHS website. The reason why I think they should be university specific is the fact that it makes the student feel like their university cares; it has made the first step so it must mean that it wants them to do well and is genuinely interested in their well-being. Being counselled should not be a gauge of failure but a normal thing everyone in need should feel encouraged to use. I know that the universities are not being idle, but I wonder what the results of a survey trying to find out if the students feel like the university empathises with them, might be like?

# References
Chang, A. (2019). The Unspoken Reality Behind the Harvard Gates. [video] Available at: https://www.youtube.com/watch?v=WzP7oDCciGI [Accessed 9 Nov. 2019].

Mathsisfun.com. (2019). Confidence Intervals. [online] Available at: https://www.mathsisfun.com/data/confidence-interval.html [Accessed 9 Nov. 2019].

Medium. (2019). Cross Validation Explained: Evaluating estimator performance.. [online] Available at: shorturl.at/kuMV7 [Accessed 9 Nov. 2019].

Soltoff, B. (2019). Cross-validation methods. [Blog] Computing for the Social Sciences. Available at: https://cfss.uchicago.edu/notes/cross-validation/ [Accessed 9 Nov. 2019].

Statistics How To. (2019). Confidence Interval: How to Find a Confidence Interval: The Easy Way! - Statistics How To. [online] Available at: https://www.statisticshowto.datasciencecentral.com/probability-and-statistics/confidence-interval/#CIProp [Accessed 9 Nov. 2019].

Statistics How To. (2019). Z-Score: Definition, Formula and Calculation - Statistics How To. [online] Available at: https://www.statisticshowto.datasciencecentral.com/probability-and-statistics/z-score/ [Accessed 9 Nov. 2019].

Statistics How To. (2019). Z-table (Right of Curve or Left) - Statistics How To. [online] Available at: https://www.statisticshowto.datasciencecentral.com/tables/z-table/ [Accessed 9 Nov. 2019].

Stulp, G. (2019). Different ways of calculating rowmeans on selected variables in a tidyverse framework - Gert Stulp. [online] Gertstulp.com. Available at: https://www.gertstulp.com/post/different-ways-of-calculating-rowmeans-on-selected-variables-in-a-tidyverse-framework/ [Accessed 9 Nov. 2019].

Wickham, H. and Grolemund, G. (2016). R for Data Science. Sebastopol, CA: O'Reilly Media.
